---
title: "Coronal Holes data features extraction"
author: "Dmitry A. Grechka"
date: "May 27, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Libraries used

```{r dependencies,message=FALSE}
library(caret)
library(ggplot2)
library(corrplot)
```

# Data sets

There are two files of sun observations for 2015 year

  * ch_2015_0193.dsv - green spectrum portion originated features
  * ch_2015_0211.dsv - red spectrum portion originated features
  
There are blue spectrum (171) images, but there are no extracted features from them

Working with ch_2015_0193.dsv fist

```{r data load}
ch_2015_0193 <- read.csv('SampleData/ch_2015_0193.dsv',
                         sep = '\t',
                         dec = ',',
                        colClasses=c("dt_record"="character")
                         )

ch_2015_0193$dt_record <- strptime(ch_2015_0193$dt_record,"%Y-%m-%dT%H:%M:%S",tz='UTC')
ch_2015_0193$ts <- as.numeric(ch_2015_0193$dt_record)
```

# Data overview

```{r data_overview}
str(ch_2015_0193)
```

# Data cleanup

There are variables that vary poorly

```{r near_zero_vars}
nzv_stats <- nearZeroVar(ch_2015_0193,saveMetrics = T)
nzv_stats[nzv_stats$nzv,]

nzv <- nearZeroVar(ch_2015_0193,saveMetrics = F)
ch_2015_0193 <- ch_2015_0193[,index2vec(nzv,vars=ncol(ch_2015_0193)) != 1]

str(ch_2015_0193)
```

#Transforms

```{r transforms}
ch_2015_0193$log_CHArea <- log(ch_2015_0193$CHArea)
```

#Correlations

```{r correlations}
numeric_vars <- colnames(ch_2015_0193[,!(names(ch_2015_0193) %in% c('fitsname','dt_record'))]) # omitting time

corM <- cor(ch_2015_0193[,names(ch_2015_0193) %in% numeric_vars],use = 'complete.obs')
varMcol<- colorRampPalette(c("red", "white", "blue"))(20)
heatmap(x = corM, col = varMcol, symm = TRUE)
#corrplot(corM,method = 'pie',order = 'hclust')
```

#Densities

```{r densities plots}

for(name in numeric_vars) {
  plot(density(ch_2015_0193[[name]],na.rm=T),main = paste('Density of',name))
}
```

# Acknowledgments

I want to thank Skobeltsyn Institute of Nuclear Physics of Moscow State University <http://swx.sinp.msu.ru> that publishes observation data on space whether and especially Lucy Mukhametdinova from SINP MSU who gave me the data in convenient CSV form