---
title: "Solar Wind observation data clean up"
author: "Dmitry A. Grechka"
date: "June 8, 2016"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Intro

We have per minute sampling of solar wind density, temperature and velocity recorded by EPAM <http://sd-www.jhuapl.edu/ACE/EPAM/> instrument of Advanced Composition Explorer (ACE) <http://sd-www.jhuapl.edu/ACE/> spacecraft.

I want to aggregate the data hourly. For each hour I want to get separate "trend"" value and noise level for each of three data variables (density, temperature, velocity). Thus I will get 

  * **density_mean**
  * **density_noise_sd**
  * **temperature_mean**
  * **temperature_noise_sd**
  * **velocity_mean**
  * **velocity_noise_sd**
  
for every hour.

# Data processing

```{r data load}
sw_2015 <- read.csv('SampleData/ace_swepam_2015.dsv',
                         sep = '\t',
                         dec = '.',
                        colClasses=c("dt_record"="character")
                         )
```

As a time axis we will use hours since 2015-01-01T00:00:00

```{r time_parsing}
base_ts_sec <- as.numeric(strptime("2015-01-01T00:00:00","%Y-%m-%dT%H:%M:%S",tz='UTC'))

dt_record <- strptime(sw_2015$dt_record,"%Y-%m-%dT%H:%M:%S",tz='UTC')
ts_sec <- as.numeric(dt_record)
sw_2015$ts <- (ts_sec - base_ts_sec)/3600;#hours

sw_2015 <- sw_2015[,names(sw_2015) != 'dt_record']

```

```{r summary}
head(sw_2015)
summary(sw_2015)
```

We will form trend by running median over the data.

```{r trend formation}
k <- 21

sw_2015$temperature.trend <- runmed(sw_2015$temperature,k)
sw_2015$velocity.trend <- runmed(sw_2015$velocity,k)
sw_2015$density.trend <- runmed(sw_2015$density,k)
```

Forming noise variables by substracting trend from original signal.

```{r noise formation}
sw_2015$temperature.noise <- sw_2015$temperature-sw_2015$temperature.trend
sw_2015$velocity.noise <- sw_2015$velocity-sw_2015$velocity.trend
sw_2015$density.noise <- sw_2015$density-sw_2015$density.trend
```

To perform per hour aggregation, I will add synthetic column containing the begining of the hour period.

```{r group column}
sw_2015$ts_h <- floor(sw_2015$ts)
```

Now we can aggreagte values within each group.

```{r aggregation}
#trend is aggregated with mean
temperature.trend.1h <- aggregate(temperature.trend ~ ts_h,data = sw_2015,mean)
velocity.trend.1h <- aggregate(velocity.trend ~ ts_h,data = sw_2015,mean)
density.trend.1h <- aggregate(density.trend ~ ts_h,data = sw_2015,mean)

#noise is aggregated with standard deviation (sd)
sd.safe <- function(x) { return(sd(x,na.rm = T))}

temperature.noise.1h <- aggregate(temperature.noise ~ ts_h,data = sw_2015,sd.safe)
velocity.noise.1h <- aggregate(velocity.noise ~ ts_h,data = sw_2015,sd.safe)
density.noise.1h <- aggregate(density.noise ~ ts_h,data = sw_2015,sd.safe)

#time is aggregared with mean (to get "centre" of aggregation time interval)
ts.1h <- aggregate(ts ~ ts_h,data = sw_2015,mean)
```

Merging (outer joining) aggregated values into single new data frame

```{r merging}
result <- merge(temperature.trend.1h,velocity.trend.1h, by = "ts_h", all = TRUE)
result <- merge(result,density.trend.1h, by = "ts_h", all = TRUE)
result <- merge(result,temperature.noise.1h, by = "ts_h", all = TRUE)
result <- merge(result,velocity.noise.1h, by = "ts_h", all = TRUE)
result <- merge(result,density.noise.1h, by = "ts_h", all = TRUE)
result <- merge(result,ts.1h, by = "ts_h", all = TRUE)

#renaming columns
result <- data.frame(
  ts=result$ts,
  density_mean=result$density.trend,
  density_noise_sd=result$density.noise,
  velocity_mean=result$velocity.trend,
  velocity_noise_sd=result$velocity.noise,
  temperature_mean=result$temperature.trend,
  temperature_noise_sd=result$temperature.noise
)

head(result)
summary(result)
```

#Visualizing the results

Visualizing first week of data

```{r visualize week}
library(ggplot2)

result_w <- result[result$ts<=168,]
raw_w <- sw_2015[sw_2015$ts<=168,]

result_m <- result[result$ts<=720,]
raw_m <- sw_2015[sw_2015$ts<=720,]

shape_scale <- c(shape1=1)
shape_scale_labels <- c(shape1='Raw observations')
fill_scale <- c(fill1='pink',fill2='darkturquoise',fill3='darkseagreen1')
fill_scale_labels <- c(fill1='Mean +/- 1 sigma noise',fill2='Mean +/- 1 sigma noise',fill3='Mean +/- 1 sigma noise')
colour_scale <- c(col1='red',col2='blue',col3='green')
colour_scale_labels <- c(col1='Mean',col2='Mean',col3='Mean')

#Velocity
p <- ggplot(aes(x=ts),data=result_w) +
  geom_point(aes(y=velocity, shape="shape1"),data=raw_w,col='darkgrey') +
  geom_ribbon(aes(ymin=velocity_mean-velocity_noise_sd,ymax=velocity_mean+velocity_noise_sd,fill ='fill1')) +
  geom_line(aes(y=velocity_mean,col='col1'),lwd=1) +
  theme_bw() +
  xlab("Time (hours since 2015-01-01T00:00:00)") +
  ylab("Solar Wind Velocity (km s-1)") +
  ggtitle("First Week of 2015 year") +
  scale_shape_manual(name="",values=shape_scale,labels=shape_scale_labels) +
  scale_fill_manual(name="",values=fill_scale,labels=fill_scale_labels) +
  scale_colour_manual(name="",values=colour_scale,labels=colour_scale_labels)
print(p)

#Density
p <- ggplot(aes(x=ts),data=result_w) +
  geom_point(aes(y=density, shape="shape1"),data=raw_w,col='darkgrey') +
  geom_ribbon(aes(ymin=density_mean-density_noise_sd,ymax=density_mean+density_noise_sd,fill ='fill2')) +
  geom_line(aes(y=density_mean,col='col2'),lwd=1) +
  theme_bw() +
  xlab("Time (hours since 2015-01-01T00:00:00)") +
  ylab("Solar Wind Density (particles cm-3)") +
  ggtitle("First Week of 2015 year") +
  scale_shape_manual(name="",values=shape_scale,labels=shape_scale_labels) +
  scale_fill_manual(name="",values=fill_scale,labels=fill_scale_labels) +
  scale_colour_manual(name="",values=colour_scale,labels=colour_scale_labels)
print(p)

#Temprature
p <- ggplot(aes(x=ts),data=result_w) +
  geom_point(aes(y=temperature, shape="shape1"),data=raw_w,col='darkgrey') +
  geom_ribbon(aes(ymin=temperature_mean-temperature_noise_sd,ymax=temperature_mean+temperature_noise_sd,fill ='fill3')) +
  geom_line(aes(y=temperature_mean,col='col3'),lwd=1) +
  theme_bw() +
  xlab("Time (hours since 2015-01-01T00:00:00)") +
  ylab("Solar Wind Temperature (?)") +
  ggtitle("First Week of 2015 year") +
  scale_shape_manual(name="",values=shape_scale,labels=shape_scale_labels) +
  scale_fill_manual(name="",values=fill_scale,labels=fill_scale_labels) +
  scale_colour_manual(name="",values=colour_scale,labels=colour_scale_labels)
print(p)

```

```{r visualize month}

result_m <- result[result$ts<=720,]

#Velocity
p <- ggplot(aes(x=ts),data=result_m) +
  geom_ribbon(aes(ymin=velocity_mean-velocity_noise_sd,ymax=velocity_mean+velocity_noise_sd,fill ='fill1')) +
  geom_line(aes(y=velocity_mean,col='col1'),lwd=0.5) +
  theme_bw() +
  xlab("Time (hours since 2015-01-01T00:00:00)") +
  ylab("Solar Wind Velocity (km s-1)") +
  ggtitle("First 30 days of 2015 year") +
  scale_fill_manual(name="",values=fill_scale,labels=fill_scale_labels) +
  scale_colour_manual(name="",values=colour_scale,labels=colour_scale_labels)
print(p)

#Density
p <- ggplot(aes(x=ts),data=result_m) +
  geom_ribbon(aes(ymin=density_mean-density_noise_sd,ymax=density_mean+density_noise_sd,fill ='fill2')) +
  geom_line(aes(y=density_mean,col='col2'),lwd=0.5) +
  theme_bw() +
  xlab("Time (hours since 2015-01-01T00:00:00)") +
  ylab("Solar Wind Density (particles cm-3)") +
  ggtitle("First 30 days of 2015 year") +
  scale_fill_manual(name="",values=fill_scale,labels=fill_scale_labels) +
  scale_colour_manual(name="",values=colour_scale,labels=colour_scale_labels)
print(p)

#Temprature
p <- ggplot(aes(x=ts),data=result_m) +
  geom_ribbon(aes(ymin=temperature_mean-temperature_noise_sd,ymax=temperature_mean+temperature_noise_sd,fill ='fill3')) +
  geom_line(aes(y=temperature_mean,col='col3'),lwd=0.5) +
  theme_bw() +
  xlab("Time (hours since 2015-01-01T00:00:00)") +
  ylab("Solar Wind Temperature (?)") +
  ggtitle("First 30 days of 2015 year") +
  scale_fill_manual(name="",values=fill_scale,labels=fill_scale_labels) +
  scale_colour_manual(name="",values=colour_scale,labels=colour_scale_labels)
print(p)
```

# Saving results

```{r outputting}
write.csv(file='ResultData/ACE_EPAM_SW_2015.csv',row.names = F,result)
```

Processing 2014 year

```{r year 2014}
sw_2014 <- read.csv('SampleData/ace_swepam_2014.dsv',
                         sep = '\t',
                         dec = '.',
                        colClasses=c("dt_record"="character")
                         )

dt_record <- strptime(sw_2014$dt_record,"%Y-%m-%dT%H:%M:%S",tz='UTC')
ts_sec <- as.numeric(dt_record)
sw_2014$ts <- (ts_sec - base_ts_sec)/3600;#hours
sw_2014 <- sw_2014[,names(sw_2014) != 'dt_record']

sw_2014$temperature.trend <- runmed(sw_2014$temperature,k)
sw_2014$velocity.trend <- runmed(sw_2014$velocity,k)
sw_2014$density.trend <- runmed(sw_2014$density,k)

sw_2014$temperature.noise <- sw_2014$temperature-sw_2014$temperature.trend
sw_2014$velocity.noise <- sw_2014$velocity-sw_2014$velocity.trend
sw_2014$density.noise <- sw_2014$density-sw_2014$density.trend

sw_2014$ts_h <- floor(sw_2014$ts)
temperature.trend.1h <- aggregate(temperature.trend ~ ts_h,data = sw_2014,mean)
velocity.trend.1h <- aggregate(velocity.trend ~ ts_h,data = sw_2014,mean)
density.trend.1h <- aggregate(density.trend ~ ts_h,data = sw_2014,mean)
temperature.noise.1h <- aggregate(temperature.noise ~ ts_h,data = sw_2014,sd.safe)
velocity.noise.1h <- aggregate(velocity.noise ~ ts_h,data = sw_2014,sd.safe)
density.noise.1h <- aggregate(density.noise ~ ts_h,data = sw_2014,sd.safe)
ts.1h <- aggregate(ts ~ ts_h,data = sw_2014,mean)
result <- merge(temperature.trend.1h,velocity.trend.1h, by = "ts_h", all = TRUE)
result <- merge(result,density.trend.1h, by = "ts_h", all = TRUE)
result <- merge(result,temperature.noise.1h, by = "ts_h", all = TRUE)
result <- merge(result,velocity.noise.1h, by = "ts_h", all = TRUE)
result <- merge(result,density.noise.1h, by = "ts_h", all = TRUE)
result <- merge(result,ts.1h, by = "ts_h", all = TRUE)

result <- data.frame(
  ts=result$ts,
  density_mean=result$density.trend,
  density_noise_sd=result$density.noise,
  velocity_mean=result$velocity.trend,
  velocity_noise_sd=result$velocity.noise,
  temperature_mean=result$temperature.trend,
  temperature_noise_sd=result$temperature.noise
)

write.csv(file='ResultData/ACE_EPAM_SW_2014.csv',row.names = F,result)
```